---
layout: topic
title: "Preprocess survey data for metrics, scientific literacy and attitude"
author: "H Qin"
date: "2019 May 31"
output: html_document
---

------------

> ## Learning Objectives
>
> * Load external tabular data from a .csv file into R.
> * Describe what an R data frame is.
> * Summarize the contents of a data frame in R.
> * Manipulate categorical data in R using factors.

------------

# Check files in the working directory 

```{r list}
list.files()
```

# Read the survey data in csv format

```{r read}
# colClass specify that all columns will be treated as characters for now. 
tb.ori = read.csv("metric_survey_data.csv", colClass=rep("character", 24))
?str
str(tb.ori); 
tb.ori$Timestamp

tb = tb.ori  #make a copy because we will modify the table. 

```

# Shorten columns names
```{r rename }
names(tb.ori)
?names
#rename the columns with shortter names for convenience 
names(tb) = c("time","gender", "age", "degree", "country", "light", "shaq", "fossil", "kilo", "mm", 
        "food","inseam", "weather","electronCharge","earlyHuman", 
        "laser", "continents", "antibiotics", "electronSize","earthCenter",
        "religiousView","dailyLife","SciOnLife", "SciEffect")
str(tb)

```


Visually check of the renamed columns.  

cbind is to combine columns. 

substr is to take a portion of string variables.  

```{r}
cbind (names(tb), substr(names(tb.ori), 1, 30))
?cbind
```


```{r}
#indexing features of R
tb[1:5, 2:3]
tb$age #what does mean? 
#tb$age[?] #try for 5th row in age

#correct some input errors
# If there is no input of 'age'
tb$age[is.na(tb$age)] = 'Do not wish to answer'
table(tb$age)
?table
# If there is no input of 'age'

tb$degree [is.na(tb$degree)] = 'Do not wish to answer'
table(tb$degree)

tb$gender[tb$gender=='']='Do not wish to answer'
table(tb$gender)

```

# Change all missing values (skipped questions) to NA
doing repetive task one by one is tedious.
we can use nested for-loops

```{r}
# dealing with missing values, add 'NA' to empty answers
# nested for-loops
for( i in 1:length(tb[, 1])) {  #outter for-loop, i for row, from 1 to the last row
  for( j in 5:length(tb[1, ])) {  #inner for-loop, j for column, from 5th to the last column
    # print( paste("i=", i, "j=", j) )
  }
}

#if there is empty cell, we assign a missing value 'NA' there
for( i in 1:length(tb[, 1])) {  #outter for-loop 
  for( j in 5:length(tb[1, ])) {  #inner for-loop
    if ( is.na(tb[i, j]) ) {
      # do nothing
    } else if (tb[i,j]=='') {
      tb[i,j] = NA
    } 
  }
}

```

# Now, we need to conver survey data in chacracters into numeric scores

# First, convert age categories into numeric values

```{r}
##### create a second table, convert character values to numerical values
tb2 = tb[ ,c(2,4,5)]  #this is the score table, empty space before comma indicate every row
head(tb2)

#calculate the average age for each category
?grep #This is not GRE prep. This is pattern match. 
# grep(pattern, x, ignore.case = FALSE, perl = FALSE, value = FALSE,
#     fixed = FALSE, useBytes = FALSE, invert = FALSE)

tb2$age = NA
tb2$age[grep("18-22", tb$age)] = 18/2 + 22/2
tb2$age[grep("23-30", tb$age)] = 23/2 + 30/2
tb2$age[grep("31-40", tb$age)] = 31/2 + 40/2
tb2$age[grep("41-50", tb$age)] = 41/2 + 50/2
tb2$age[grep("51-55", tb$age)] = 51/2 + 55/2
tb2$age[grep("56-60", tb$age)] = 56/2 + 60/2
#> grep("56-60", tb$age)
# [1]   6  12  84  88 104 112 129 130 131 132 133 156 162 163 185 216 219

tb2$age[grep("More than 60 years", tb$age)] = 65
```

Check the age responses
```{r}
table(tb$age) #table is a very useful function (command) for tabulation
table(tb2$age)
summary(tb2$age)
```

# Visualize the age values

```{r}
table(tb2$age, tb2$gender)
boxplot( tb2$age ~ tb2$gender)

#histogram of age
hist(tb2$age)

```

# Convert country responses into values

```{r}
table( tb$country )  #All the inputed 'countries'
tb2$country = 0  #for non-USA countries
tb2$country[tb$country=='United States'] = 1
table( tb2$country )

#have a look at some entries
head(tb2)

#double-check the columns
names(tb2)
```

# The survey contains by 3 categories of questions
1) Metric proficiency
2) Scientific literacy
3) Attitude toward science

Tocalculate the score of each categoriy separately, we need to identify these columns. 

```{r define-3-categories}
### Here are the columns for the 3 categories
metrics = c("shaq", "kilo", "mm", "inseam", "weather")
sciLiteracy = c("light", "fossil", "food", "electronCharge", 
                "earlyHuman", "laser", "continents", "antibiotics",
                "electronSize", "earthCenter")
sciAttitude = c("religiousView", "dailyLife", "SciOnLife", "SciEffect")
```

# Calculate the metric-proficiency scores

```{r metric-proficiency-score}

tb2$shaq = 0
tb2$shaq[ tb$shaq=='Yes' ] = 1
tb2$shaq[ tb$shaq=='No' ] = 0
table(tb2$shaq)

tb2$kilo = 0
tb2$kilo[ tb$kilo=='1000 x' ] = 1
table(tb2$kilo)

tb2$mm=0
tb2$mm[ tb$mm==0.145 ] = 1
table(tb2$mm)
table(tb$mm)

tb2$inseam = 0
tb2$inseam[tb$inseam=="This person is short"] = 1
tb2$inseam[tb$inseam=="This person is tall"] = 0
table(tb2$inseam)

tb2$weather = 0
tb2$weather[tb$weather=="A Short sleeve shirt"] = 1
#tb2$weather[tb$weather=="A winter coat"] = 0
#tb2$weather[tb$weather=="A light jacket"] = 0
table(tb$weather)
table(tb2$weather)


```

# Summarize the metric proficiency score by rows. 

```{r}
# metrics = c("shaq", "kilo", "mm", "inseam", "weather")
#metric total score
tb2$metric = apply( tb2[, metrics], MARGIN=1, FUN=sum )
hist(tb2$metric, br=4)
```

# Calcualte the science attitude scores

```{r}

#sciAttitude = c("religiousView", "dailyLife", "SciOnLife", "SciEffect")
# "My religious views are more important than scientific views
tb2$religiousView = 0
tb2$religiousView[grep("No", tb$religiousView)] = 1
tb2$religiousView[grep("Yes", tb$religiousView)] = 0
table(tb2$religiousView)
table(tb$religiousView)

# "For me, in my daily life, it is not important to know about science"
tb2$dailyLife = 0
tb2$dailyLife[ tb$dailyLife=='TRUE' ] = 0
tb2$dailyLife[ tb$dailyLife=='FALSE' ] = 1
table(tb2$dailyLife)

# "Science and technology are making our lives healthiers, easiers and more comfortable."
tb2$SciOnLife = 0
tb2$SciOnLife[ tb$SciOnLife=='TRUE' ] = 1
tb2$SciOnLife[ tb$SciOnLife=='FALSE' ] = 0
table(tb2$SciOnLife)

# "The benefits of sciences are greaters than any harmful effects that it may have."
tb2$SciEffect = 0
tb2$SciEffect[ tb$SciEffect=='TRUE' ] = 1
tb2$SciEffect[ tb$SciEffect=='FALSE' ] = 0
table( tb2$SciEffect )

#sciAttitude = c("religiousView", "dailyLife", "SciOnLife", "SciEffect")
#Attitude total score
tb2$SciAttitude = apply( tb2[, sciAttitude], MARGIN=1, FUN=sum)
hist(tb2$SciAttitude, br=20)
```


#Calculate scientific literacy

```{r}
#sciLiteracy = c("light", "fossil", "food", "electronCharge", 
#                "earlyHuman", "laser", "continents", "antibiotics", "electronSize", "earthCenter")
tb2$light = 0
tb2$light[ tb$light=='TRUE' ] =1
tb2$light[ tb$light=='Wrong' ] =0
table(tb$light)
table(tb2$light)

tb2$fossil = 0
tb2$fossil[ tb$fossil=='6 million and 5 years old' ] = 0
tb2$fossil[grep('Still', tb$fossil)] = 1;
table(tb$fossil)
table(tb2$fossil)

tb2$food = 0
tb2$food[ tb$food=='Dis-agree' ] = 1
tb2$food[grep('Agree', tb$food)] = 0; 
table(tb$food)
table(tb2$food)

tb2$electronCharge = 0
tb2$electronCharge[grep('Positive', tb$electronCharge)] = 1; 
table(tb$electronCharge)
table(tb2$electronCharge)

tb2$earlyHuman = 0
tb2$earlyHuman[grep('TRUE', tb$earlyHuman)] = 0; 
tb2$earlyHuman[grep('FALSE', tb$earlyHuman)] = 1; 
table(tb$earlyHuman)
table(tb2$earlyHuman)

tb2$earlyHuman = 0
tb2$earlyHuman[grep('TRUE', tb$earlyHuman)] = 0; 
tb2$earlyHuman[grep('FALSE', tb$earlyHuman)] = 1; 
table(tb$earlyHuman)
table(tb2$earlyHuman)

tb2$laser = 0
tb2$laser[grep('TRUE', tb$laser)] = 0; 
tb2$laser[grep('FALSE', tb$laser)] = 1; 
table(tb$laser)
table(tb2$laser)

tb2$continents = 0
tb2$continents[grep('TRUE', tb$continents)] = 1; 
tb2$continents[grep('FALSE', tb$continents)] = 0; 
table(tb$continents)
table(tb2$continents)

tb2$antibiotics = 0
tb2$antibiotics[grep('TRUE', tb$antibiotics)] = 0; 
tb2$antibiotics[grep('FALSE', tb$antibiotics)] = 1; 
table(tb$antibiotics)
table(tb2$antibiotics)

tb2$electronSize = 0
tb2$electronSize[grep('True', tb$electronSize)] = 1; 
tb2$electronSize[grep('FALSE', tb$electronSize)] = 0; 
table(tb$electronSize)
table(tb2$electronSize)

tb2$earthCenter = 0
tb2$earthCenter[grep('TRUE', tb$earthCenter)] = 1; 
tb2$earthCenter[grep('FALSE', tb$earthCenter)] = 0; 
table(tb$earthCenter)
table(tb2$earthCenter)

#sciLiteracy = c("light", "fossil", "food", "electronCharge", 
#                "earlyHuman", "laser", "continents", "antibiotics", "electronSize", "earthCenter")

tb2$SciLitScore = apply( tb2[, sciLiteracy], MARGIN=1, FUN=sum ) #by row
hist(tb2$SciLitScore, br=20)
```

```{r}
str(tb2)
```

# Output the 'cleaned' data to a csv file

```{r}
tb3 = tb2[, c("gender", "age", "country", "degree", "metric", "SciAttitude", "SciLitScore")]
head(tb3)
write.csv(tb3, file = "metric-attitude-literacy.csv", row.names = FALSE, quote=TRUE)
```
