---
title: "Analysis of survey data for metric proficiency, scientific literacy and attitude"
author: "H Qin"
date: "5/18/2020"
output:
  html_document: default
  pdf_document: default
layout: topic
---

------------

> ## Learning Objectives
>
> * Load external tabular data from a .csv file into R.
> * Describe what an R data frame is.
> * Summarize the contents of a data frame in R.
> * Manipulate categorical data in R using factors.

------------

# Check files in the working directory 

```{r list, echo=FALSE}
rm( list=ls()) #clean up worksapce
list.files()
```

# Read the survey data in csv format

```{r read}
tb = read.csv("metric-attitude-literacy.csv", stringsAsFactors = FALSE)
```

What is a data frame? 

Looking for helps
```{r help}
?str
help(str);
str(tb);
```

Indexing an element in a dataframe

```{r indexing}
#indexing features of R
tb[1:5, 2:3]
tb$age #what does mean? 
#tb$age[?] #try for 5th row in age

# what is the 5th and 9th rows in age and SciLitScore: 
tb[c(5,9), c(7,2)]
```

```{r indexing-by-c}
tb[c(5,9), c('age', 'SciLitScore')]
```

# tabuluate the values of 'age'

```{r}
table(tb$age)
```


Overview of the data

```{r summary}
summary(tb)
```
Exercise: Find out help information of summary(). 

```{r look-at-data}
head(tb)
```

Look at first 2 rows of columns 2 and 3
```{r}
head( tb[, 2:3], n=2)
```

# The survey provide results for three types of questions
1) Metric proficiency
2) Scientific literacy
3) Attitude toward science

What does the 'country' coded for? 

```{r}
table(tb$country)
```
So, USA is 1, other countries are coded as zeros. 

# Is gender associated with metric proficiency? 

```{r}
boxplot( tb$metric ~ tb$gender, ylab="metric proficiency" )
```

```{r}
table(tb$metric, tb$gender )
```



How to pick a subset of data? 
Pick "Male" metric proficiency data? 
```{r}
tb$gender == 'Male'
tb$metric[tb$gender=='Male']
tb$metric[tb$gender=='Female']
```

What is t-test? 

```{r t-test}
t.test(tb$metric[tb$gender=='Female'], tb$metric[tb$gender=='Male'], alternative = 'less')
# Does this mean that females are more uncomfortable with metric usage?
```

# What does p-value mean? 

```{r boxpliot}
# Female participants tend to be younger
boxplot( tb$age ~ tb$gender, ylab='age')

# More female participants with Bachelor degrees
table( tb$gender, tb$degree )

boxplot( tb$metric ~ tb$degree, ylab='metric proficiency')

# how age in boxplot
# ???
  
m1 = lm( tb$metric ~ tb$gender )
summary(m1)
#summary( lm( tb$metric ~ tb$gender ) )

m2 = lm( tb$metric ~ tb$age )
summary(m2)

m3 = lm( tb$metric ~ tb$degree )
summary(m3)

m4 = lm( tb$metric~ tb$gender + tb$age + tb$degree )
summary(m4)
```


# Caudal regression analyses

```{r}
#########################
#summary(tb); str(tb)

#remove rows with missing age from analysis. Missing age can cause bugs in anova model comparisons. 
tb = tb[!is.na(tb$age), ] 

summary(tb)
str(tb); 

pairs(tb[, c("age", "metric", "SciLitScore", "SciAttitude")])

```


```{r causal-analsyis}
summary(lm(tb$SciLitScore ~ tb$metric )) #significant
summary(lm(tb$SciAttitude ~ tb$metric )) #significant
summary(lm(tb$SciAttitude ~ tb$SciLitScore )) #significant
summary(lm(tb$SciAttitude ~ tb$age )) #significant
summary(lm(tb$metric ~ tb$age )) #significant

summary(lm(tb$SciAttitude ~ tb$SciLitScore + tb$metric )) #significant
## metric -> SciAttitude and SciLitScore


summary(lm(tb$SciAttitude ~ tb$metric + tb$age + tb$gender + tb$country  )) #only metric is significant
summary(lm(tb$SciLitScore ~ tb$metric + tb$age + tb$gender + tb$country  )) #only metric is significant

summary(lm(tb$SciLitScore ~ tb$country)) #p=0.0009, but seems due to metric?  
summary(lm(tb$SciLitScore ~ tb$metric + tb$country  )) #only metric is significant

summary(lm(tb$SciAttitude ~ tb$country)) #p=0.0127, but seems due to metric?  
summary(lm(tb$SciAttitude ~ tb$country + tb$metric)) #country not significant when controled for metric 


plot( tb$SciLitScore ~ jitter(tb$metric), xlab='Metric Proficiency', ylab='Scientific Literacy', ylim=c(2,10) )
m1 = lm(tb$SciLitScore ~ tb$metric )
abline(m1, col='red')
summary(m1)
text(2, 2.5, "SciLit ~ Metric, R2=0.26, p<0.001", col="red", pos=4)
#abline(m2, col='blue')
summary(m2)
m2 = lm(tb$SciLitScore ~ tb$metric + tb$age)
anova(m1, m2)
m3 = lm(tb$SciLitScore ~ tb$metric + tb$age + tb$gender)
summary(m3)
anova(m2,m3)
m4 = lm(tb$SciLitScore ~ tb$metric + tb$age + tb$country)
anova(m2, m4)
#text(2, 2, "SciLit ~ Metric + Age, R2=0.29, p=2.8E-14", col="blue", pos=4)

plot( tb$SciAttitude ~ jitter(tb$metric), ylim=c(0.5,4), xlab='Metric Proficiency', ylab='Attitude toward Science' )
m1 = lm( tb$SciAttitude ~ tb$metric )
m2 = lm( tb$SciAttitude ~ tb$metric + tb$age )
abline(m1, col='red')
abline(m2, col='blue')
summary(m1)
summary(m2)
anova(m1, m2)
text(2, 0.9, "SciAttitude ~ Metric , R2=0.18, p=1.0E-9", col="red", pos=4)
text(2, 0.7, "SciAttitude ~ Metric + Age, R2=0.24, p=4.7E-12", col="blue", pos=4)

plot( tb$SciAttitude ~ jitter(tb$age), ylab='Attitude toward Science', xlab='Age')
m2 = lm( tb$SciAttitude ~ tb$age + tb$metric)
abline(m2, col='blue')
text(30, 1.7, "SciAttitude ~ Metric + Age, R2=0.24, p=4.7E-12", col="blue", pos=4)

summary(lm(tb$SciAttitude ~ tb$metric + tb$age + tb$gender + tb$country  )) #age is signicant!!!

#but this might be a bias in the sample
# 1) there is many faculty
# 2) people took the sample may be interested in the metric and science at the first place?!
summary(lm(tb$SciAttitude ~ tb$metric + tb$age + tb$gender + tb$country + tb$degree  )) #age is signicant!!!

summary(lm(tb$SciAttitude ~ tb$SciLitScore))
summary(lm(tb$SciAttitude ~ tb$SciLitScore + tb$metric))

boxplot( tb$SciAttitude ~ tb$country )
boxplot( tb$SciLitScore ~ tb$country )
boxplot( tb$metric ~ tb$country )

boxplot( tb$SciLitScore ~ tb$age )
boxplot( tb$SciAttitude ~ tb$age )
boxplot( tb$metric ~ tb$age )

###########
# remove phD from the samples
#
summary(tb[, 1:5])
tb3 = tb[ - grep('Ph.D.', tb$degree)  , ]
summary(tb3)
summary(lm(tb3$SciAttitude ~ tb3$metric + tb3$age + tb3$gender + tb3$country + tb3$degree  )) 
#age is still signicant after PhD are removed from the sample

```

