---
title: "Simple statistics, 2020 Election Reults"
author: "H Qin"
date: "5/18/2021"
output:
  html_document: 
    toc: true
    toc_float: true
---

# Read a data set

```{r}
rm(list=ls()) #clearn my workspace
library(tidyverse)
Election20df = read_csv("https://raw.githubusercontent.com/tonmcg/US_County_Level_Election_Results_08-20/master/2020_US_County_Level_Presidential_Results.csv")
head(Election20df)

```

# Question: What is the total number of votes? 

```{r}
Election20df %>% select ( total_votes ) %>% sum()

sum( Election20df$total_votes ) /1E6
```

# Question: What are the total number votes for GOP or DEM? 

```{r}
Election20df %>% select ( votes_gop ) %>% sum()
Election20df %>% select ( votes_dem ) %>% sum()

```

# Look at California
```{r}
mystate = "California"
Californiadf <- 
  Election20df %>% filter( state_name == mystate) %>%  arrange( per_point_diff)

names( Californiadf )[8] = "percentage_for_GOP"

ggplot(Californiadf, aes(percentage_for_GOP)) +  geom_histogram()
#summary(Californiadf$per_gop)
```

# Results by States? 

```{r}
Statedf <- 
  Election20df %>% select( state_name, votes_gop, votes_dem,  total_votes ) %>% group_by( state_name ) %>% summarise_if( is.numeric, sum)

Statedf$percentage_for_GOP = Statedf$votes_gop / Statedf$total_votes

ggplot(Statedf, aes(percentage_for_GOP)) +  geom_histogram()
```

# Question: Find out which state has the nearly 95% DEM voting percentage? 

There are many ways to do this. 

```{r}
Statedf %>% filter( percentage_for_GOP < 0.1 )
```

# Average 

```{r}
mean( Statedf$percentage_for_GOP)
```

```{r}
quantile( Statedf$percentage_for_GOP )
```

# Label the states to deep red, red, swing, blue, deep blue

Reference: 
https://stackoverflow.com/questions/21050021/create-category-based-on-range-in-r
```{r}
groups =  cut( Statedf$percentage_for_GOP, c(0, 0.4, 0.47, 0.53, 0.6, 1) )
levels(groups) =                   c("deepblue", "blue", "swing", "red", "deepred")
Statedf$groups = groups
```

# add cencus data

```{r}
Censusdf = read_csv("https://raw.githubusercontent.com/hongqin/USA-census-county-level/main/USA-County-level-census-2010-2019.csv")
head(Censusdf)
```

Some states, such as Alaska, election results are not based on counties. 

```{r}
Election20df$Location = paste( Election20df$county_name, Election20df$state_name, sep=", " )
Election20df$Location %in% Censusdf$Location
```


```{r}
EleCen.df = merge( Election20df, Censusdf, by="Location")
```

```{r}
Statedf2 <- EleCen.df %>% select( state_name, votes_gop, votes_dem,  total_votes, '2019' ) %>% group_by( state_name ) %>% summarise_if( is.numeric, sum)

head(Statedf2)
```

```{r}
names( Statedf2)[5] = "population"
```

```{r}
Statedf$population = Statedf2$population[match( Statedf$state_name , Statedf2$state_name ) ]
```

## any correlation with voting and population? 

```{r}
model1 = lm( Statedf$percentage_for_GOP ~ Statedf$population)
summary(model1)
```


```{r}
model2 = lm(  Statedf$population ~ Statedf$groups)
summary(model2)
ggplot( Statedf, aes(x=groups, y=population)) + geom_point()
```

# Add  state areas

```{r}
StateArea = read_csv("https://raw.githubusercontent.com/hongqin/data-USstates/master/state-areas.csv")
names( StateArea) = c("state_name", "area")
```

## merge different data sets

```{r}
Statedf$area = StateArea$area[ match( Statedf$state_name  , StateArea$state_name ) ]
Statedf$pop_density = Statedf$population / Statedf$area

model = lm( Statedf$percentage_for_GOP ~ Statedf$pop_density)
summary(model)
```


```{r}
ggplot( Statedf, aes(x=pop_density, y=percentage_for_GOP)) + geom_point()
```

```{r}
Statedf3 <- Statedf %>% filter( percentage_for_GOP > 0.1)

ggplot( Statedf3, aes(x=pop_density, y=percentage_for_GOP)) + 
  geom_point() + 
  geom_smooth(method='lm')
 
```

```{r}
summary(lm(Statedf3$percentage_for_GOP~ Statedf3$pop_density))
```

## voting and population density

```{r}
ggplot( Statedf3, aes(x=groups, y=pop_density)) + geom_point()
```

```{r}
ggplot( Statedf3, aes(x=groups, y=pop_density)) + geom_boxplot()
```

# t-test 

```{r}
deepred_pop_densities <- 
    Statedf3 %>% filter( groups == "deepred") %>% select( pop_density)

deepblue_pop_densities <- 
    Statedf3 %>% filter( groups == "deepblue") %>% select( pop_density)

t.test( deepblue_pop_densities, deepred_pop_densities, alternative = "greater")
```
```{r}
ggplot(Statedf, aes(x=state_name, y=percentage_for_GOP)) + geom_bar(stat='identity', width=.5)+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
ggplot(Statedf, aes(x=reorder(state_name, percentage_for_GOP), y=percentage_for_GOP)) + geom_bar(stat='identity', width=.5)+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

