---
title: "Frasher_fig3A"
author: "H Qin"
date: "10/5/2017"
output: html_document
---

```{r}
rm(list=ls())
set.seed(2017)
source('network.r')
```

```{r}
#input evolutionary distances
Kdata = read.csv( "Sce.Spa.KaKs.csv");

#input protein interaction pairs
pairs = read.csv("pairs.csv");
pairs$ORF1 = as.character( pairs$ORF1 );
pairs$ORF2 = as.character( pairs$ORF2 );

#input growth fitness
fit = read.csv("growth.fitness.hom.csv");
fit$orf = as.character( fit$orf ); 
# YPD is the fitness data that can be used for Figure 3B. 

```

# Fig 3A. Use permutation to calculate p-value of Ka difference
Use permutation to calculate p-value of Ka difference in interacting pairs;
This is equivalent to Fig 3A in Fraser 02, Science


First, define a function to calculate K difference
```{r}
 # First, define a function to calculate Ka difference in pairs of proteins
 diff.K = function( inpairs ) {
   inpairs$K1= Kdata$Ka[match(inpairs$ORF1, Kdata$orfname)];
   inpairs$K2= Kdata$Ka[match(inpairs$ORF2, Kdata$orfname)];
   ret = mean( abs( inpairs$K1 - inpairs$K2 ), na.rm=T );
 } 
 
 # calculate the observed difference in Ka
 diff.K.obs = diff.K ( pairs );
 paste( "Observed deltaK = ", diff.K.obs); 
```

```{r}

 #permutation of pairs, and their difference in Ka
 Nsims = 10; #number of permutations
 permutated.diff.K = numeric( Nsims ); #empty vector to store calculations

 # Now do N simulations
 for( i in 1:Nsims ) {
  
   new.pairs = ms02_singlerun(pairs ) #generate a new MS02 random network
   new.pairs = new.pairs[,1:2] #reformating into two-columns
   names(new.pairs) = c("ORF1", "ORF2")

   #calculate delta.K for one random network
   permutated.diff.K[i] = diff.K( new.pairs ); #for null distribution of deltaK
 }

head(permutated.diff.K)
hist(permutated.diff.K)

 # calulate p-value
 sub = permutated.diff.K[ permutated.diff.K <= diff.K.obs ]
 p.K = length(sub) / length(permutated.diff.K);
 paste("p-value of observed deltaK =", p.K);
  
 # generate Figure 3A
 hist( permutated.diff.K, xlim=c(0.025, 0.035), br=20 );
 arrows( diff.K.obs, sqrt(Nsims), diff.K.obs, 0.5, col="red" );
 text( diff.K.obs, sqrt(Nsims)*1.05, "obs");
```

# Quiz/ Assignment. Regenerate Figure 3B. 

Use permutation to find out p-value of the difference in fitness (YPD) of interacting pairs. This is equivalent to Figure 3B in Fraser 02, Science. 
