---
title: "Learning R through GISAID metadata analysis"
author: "H Qin"
date: "5/17/2021"
output:
  html_document: 
    toc: true
    toc_float: true
---


# Load GISAID meta data


------------

> ## Learning Objectives
>
> * Load external tabular data from a Excel file into R.
> * Describe what an R data frame is.
> * Explore the contents of a data frame in R.

------------

# Load the libraries

```{r list, echo=FALSE}
rm( list=ls()) #clean up worksapce
library(tidyverse)
library(lubridate)
library(readxl)
```

# Load the data

```{r read}
tb = read_excel('gisaid_monthly_submissions_global_2021-05-12-0958.xlsx')
```

#  What is a data frame? 

```{r}
str(tb)
```

Looking for helps
```{r help}
?str
help(str);
head(tb)
```

# Indexing items in a dataframe, using square brackets [ ], before comma are row, after comma are columns. 


```{r}
#indexing features of R
tb[1:5, 2:4]
```
```{r}
tb[3, 4]
```

```{r indexing}
# what is the 5th and 9th rows in age and SciLitScore: 
tb[c(3, 4, 7), c(4,5)]
```

```{r indexing-by-c}
tb[c(3,4), c( '03/2020', "04/2020")]
```

# Look at a portion of the data frame
```{r look-at-data}
tail(tb)
```

Look at first 2 rows of columns 2 and 3
```{r}
head( tb[, 2:3], n=2)
```

# How many columns? 

```{r}
# how many columns? 
length(tb[1,])
print(paste( "There are ", length(tb[1,]), " columns"))
```

## we can use $ to access the columns in data frame

```{r}
tb$`01/2020`
```

# Take only the values
```{r}
tb1 = t(tb[, 2: length(tb[1,])])
tb1 = data.frame(tb1)
names(tb1) = t(tb[,1]) #use contry names as columns names in tb1
head(tb1)
```

# Convert strings into Date type
```{r}
library(zoo)

```

```{r}
tb1$date = as.yearmon( row.names(tb1),  "%m/%Y")
ggplot( tb1, aes(x=date, y=USA)) + geom_point()
```

```{r}
names(tb1)
```

# We can use $ to take a spcifiied column

```{r}
tb1$USA
```

# tidyverse, select columns (variables)

```{r}
tb1 %>% select (USA)
```

# tidyverse, filter by rows
```{r}
tb1 %>% filter( USA > 100 & USA < 5000 ) %>% select (USA, China)
tb1[, c("USA", "China")]
```
```{r}
tb1$"monthly total:"
```

```{r}
tb1$USA_percentage = tb1$USA / tb1$"monthly total:"
tb1$USA_percentage
```

# total submission 
```{r}
sum(tb1$"monthly total:") / 1E6
```

